```{r}
rm(list=ls())
library('here')
library('dplyr')
library('ggplot2')
library("aTSA")
library("lubridate")
library("readr")
library("data.table")
library("tidyr")
require("gridExtra")
library("seqICP")
library("lmtest")
library("roll")
library("vars")
```

```{r}
target <- "ldEXME"

data <- read_csv(here('src', 'new_studies', 'processedData', 'monetary-policy-processed.csv'))  
data$...1 <- NULL

X <- data %>%
  as.data.table() %>%
  mutate(date=ymd(date)) %>%
  drop_na() %>%
  filter((date >= ymd('1999-07-01')) & (date <= ymd('2017-01-01'))) %>%
  select(-sym(target))
y <- data %>%
  as.data.table() %>%
  mutate(date=ymd(date)) %>%
  drop_na() %>%
  filter((date >= ymd('1999-07-01')) & (date <= ymd('2017-01-01'))) %>%
  select(date, sym(target))

data_lag <- data %>%
  select(-date) %>%
  as.data.table() %>%
  shift(1) %>%
  as.data.table() %>%
  setnames(paste0(names(data)[-grep("date", names(data))], "_lag1")) %>%
  mutate(date=data$date) %>%
  select(date, everything())
data_lag <- merge(data, data_lag, by='date') %>% drop_na()
```

```{r}
tail(y)
```

```{r}
tail(X)
```

```{r}
tail(data_lag)
```


```{r}
ggplot(y %>% melt(., id.vars = 'date'), aes(x = date, y = value, color = variable)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'CHF target variable', x = 'Time', y = 'Target')
```

```{r}
melt_X <- melt(X, id.vars = 'date')
ggplot(melt_X, aes(x = date, y = value, color = variable)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'CHF features', x = 'Time', y = 'Feature')
```

```{r}
colnames(X)
```

## 1. Testing Stationarity

On the application of Pfister et al. (2018) the authors seek to find instantaneous monthly causal predictors of the returns of EURCHF exchange rate.

To do this the authors build a linear model of the EURCHF exchange rate as a function of lags of both the target and the features.

The argument of the authors is based on testing invariance induced by:

> 1. Changes in variance and mean of the residuals
> 2. Changes in the coefficients of the DGP`

```{r}
formula_X <- paste0("`", paste(names(data_lag)[-grep(paste0(target, "|date"), names(data))], collapse="` + `"), "`")
fomula <- paste(paste(target, "~", formula_X), " -1")
model <- lm(fomula, data=data_lag %>% select(-date))
residual_df <- residuals(model) %>%
  as.data.table() %>%
  setnames('residual') %>%
  mutate(date=data_lag$date)

# ggplot residuals
ggplot(residual_df, aes(x = date, y = residual)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Residuals of the linear model', x = 'Time', y = 'Residuals')
```

### (a). Hypothesis testing variance-shift

ACF

> 1. Slow decay of the ACF indicates non-stationarity
> 2. This would imply that the autocovariance of the residuals is not constant over time

```{r}
acf(residual_df %>%
      select(residual), lag.max = 20, main = "Residuals ACF")
```

Goldfeld-Quandt test

$$
\text{GQ} = \frac{s_1}{s_2} \approx \frac{\hat{u}^{\intercal}_1\hat{u}_2}{\hat{u}^{\intercal}_1\hat{u}_2}
$$
$$
H_0: \text{Variance does not increase from segment 1 to 2}
$$

> 1. The Goldfeld-Quandt test is used to test for heteroscedasticity
> 2. The test is based on the assumption that the variance of the residuals is constant over time
> 3. The test requires a definition of blocks
> 4. The test assumes the existence of two blocks

```{r}
gqtest(model, fraction = 0.1)
```
```{r}
gqtest(model, fraction = 0.05)
```

Whites test (??)

### (b). Hypothesis testing mean-shift

Rolling mean of residuals

> 1. The rolling mean of the residuals should be constant over time

```{r}
residual_df <- residual_df %>%
  mutate(rolling_mean = rollmean(residual, k=24, fill=NA)) %>%
  drop_na()
ggplot(residual_df, aes(x = date, y = residual)) +
  geom_line() +
  geom_line(aes(x = date, y = rolling_mean), color = 'red') +
  theme_minimal() +
  labs(title = 'Rolling mean of residuals', x = 'Time', y = 'Residuals')
```

Time-varying coefficients

> 1. The time-varying coefficients of the linear model should be constant over time

```{r}
# rolling regression
rolling_lm <- roll_lm(x = data_lag %>% select(sym(target)) %>% as.matrix(),
                      y = data_lag %>% select(-sym(target), -date) %>% as.matrix(),
                      width = as.integer(12 * 5),
                      intercept = FALSE,
                      online = TRUE)
rolling_coefs_df <- rolling_lm$coefficients %>%
  as.data.table() %>%
  mutate(date=data_lag$date) %>%
  select(date, everything()) %>%
  drop_na()

# scale the coefficients
scale_rolling_coefs_df <- rolling_coefs_df %>%
  mutate_at(vars(-date), ~(. - mean(.)) / sd(.))
melt_scale_rolling_coefs <- melt(scale_rolling_coefs_df, id.vars = 'date')

ggplot(melt_scale_rolling_coefs, aes(x = date, y = value, color = variable)) +
  geom_line() +
  theme_minimal() +
  labs(title = 'Time-varying coefficients', x = 'Time', y = 'Coefficients')
```

## 2) seqICP

```{r}
# select optimal lag
lag_selection <- VARselect(data %>% select(-date), lag.max = 6, type = "const")
p <- as.numeric(lag_selection$selection[1] )

# pvalue for the hypothesis testing
pvalue <- 0.05
```


```{r, warning=FALSE}
res_variance <- seqICP(X %>% select(-date) %>% as.matrix(),
                       y %>% select(sym(target)) %>% as.matrix(),
                       test="variance",
                       par.test=list(alpha=pvalue, # significance level of the hypothesis test
                                     B=1000, # num of MC samples used to approximate the null dist
                                     link=sum, # how to compare pairwise test stats
                                     grid=c(0,70,140,nrow(X)), # grid points to build envs
                                     complements=TRUE # if TRUE, envs should be compared against complements
                                     ),
                       model="ar",
                       par.model=list(pknown=TRUE,p=p),
                       stopIfEmpty=FALSE,
                       silent=TRUE)
summary(res_variance)
```

```{r, warning=FALSE}
res_variance <- seqICP(X %>% select(-date) %>% as.matrix(),
                       y %>% select(sym(target)) %>% as.matrix(),
                       test="variance",
                       par.test=list(alpha=pvalue, # significance level of the hypothesis test
                                     B=1000, # num of MC samples used to approximate the null dist
                                     link=sum, # how to compare pairwise test stats
                                     grid=c(0,70,140,nrow(X)), # grid points to build envs
                                     complements=TRUE # if TRUE, envs should be compared against complements
                                     ),
                       model="ar",
                       par.model=list(pknown=TRUE,p=1),
                       stopIfEmpty=FALSE,
                       silent=TRUE)
summary(res_variance)
```


