---
title: "Time Series Properties using 'gratis' package"
output:
---


```{r}
rm(list=ls())
library("gratis")
library("forecast")
library("nortsTest")
library("seqICP")
library("data.table")
library("dplyr")
```

## Goal

The main goal of this notebook is to evaluate the usefulness of the gratis package to help us simulate time series with a range of different properties. 

Our dataset consists of montlhy returns for a range of ETFs and FRED monthly time series features of macro and financial variables. To evaluate the usefulness of gratis in this context, we must first know the stylized facts that typically describe financial and economic time series.



## Description

We will use the package called "gratis" to simulate the time series processes. This package uses Gaussian mixture autoregressive (MAR) models to simulate non-gaussian and non-linear time series.

Below we sample two time series by controlling for the following characteristics:

> 1) Trend
  params: trend
  des: measures the trend coefficient (intercept) of a STL decomposition
  domain: $[0, 1)$ 
> 2) Linearity
  params: linearity
  des: measures the linearity of a time series calculated based on the coefficients of an orthogonal
       quadratic regression.
  domain: $(-\infty, \infty)$
> 3) Autocorrelation
  params: x_acf1, x_acf10
  des: x_acf1 set the value of the first autocorrelation coefficiente, whereas x_acf10 set the value of
       the first 10 (sum of their values).
  domain: $(-1, 1)$ or $(0, \infty)$
> 4) Conditional Heteroskedasticity
  params: arch_acf
  des: arch_acf is the the sum of squares of the first 12 autocorrelations of the square of the time series.
  domain: $(0, \infty)$

```{r}
x <- generate_ts_with_target(n = 5,
                             ts.length = 200,
                             freq = 1,
                             seasonal = 0,
                             features = c("tsfeatures", "heterogeneity"),
                             selected.features = c("trend", "linearity", "x_acf1", "garch_acf"),
                             target = c(0, 0, 0, 0),
                             parallel = FALSE)
```
## Checking Time Series Properties

We will test for the above time series characteristics using the following tests:

> 1) ACF and PACF plots: This will be helpful to check for any evidence of ARMA components on the original time series. Furthremore, we can also run ACF and PACF for the squares of the time series to check for arch ancd garch effects.
> 2) 


### 1) Checking for ARMA components

```{r}
for (i in 1:dim(x)[2]){
  ggtsdisplay(x[, i])
}
```

### 2) Checking for ARCH/GARCH effects

```{r}
for (i in 1:dim(x)[2]){
  ggtsdisplay(x[, i]^2)
}
```

```{r}
for (i in 1:dim(x)[2]){
  print(shapiro.test(x[, i])) # null of normality
}
```

```{r}
# parameters
d <- dim(x)[2]
prob <- 0.5

# define adjacency matrix
entries <- rbinom(d * d, size = 1, prob = prob)
adjacency_matrix <- matrix(data = entries, nrow = d, ncol = d)

# define coefficient values
entries <- rnorm(d * d, mean = 0, sd = 0.5)
coef_matrix <- matrix(data = entries, nrow = d, ncol = d)

# add constraint of no cycles
c1 <- matrix(1, nrow = d, ncol = d)
diag(c1) <- 0

# design matrix for th DGP
B <- adjacency_matrix * coef_matrix * c1

# target as a linear transformation
x <- as.matrix(x)
y <- (x %*% B[,1] + rnorm(d * d, mean = 0, sd = 0.5)) %>% as.data.table()

# take out y from x
X <- x[,2:dim(x)[2]] %>% as.data.table()

ggtsdisplay(y)
ggtsdisplay(y^2)
shapiro.test(y %>% as.matrix()) # null of normality
B
```


```{r}
summary(lm(V1 ~ `Series 2` + `Series 3` + `Series 4` + `Series 5` -1, data = cbind(y, X) %>% as.data.table()))
```

```{r}
seqICP(X=X,
       Y=y,
       test = "hsic",
       model = "ar",
       stopIfEmpty = FALSE,
       silent = TRUE)
```








