---
title: "Test seqICP on ETFs data - Daniel"
output:
---


```{r}
rm(list=ls())
library("dplyr")
library("seqICP")
library("readr")
library("here")
library("data.table")

inputs_path <- file.path(here(), "src", "data", "inputs")
```

First, lets load the dataset:

```{r}
df <- read_csv(file = file.path(inputs_path, "monthly_ret_etfs_macro_tranfs.csv"))
head(df)
```
```{r}
dim(df)
```

Below, we list the hypothesis of the seqICP:

> 1) The relationship of yt on Xt is linear on the parameters (linearity)
> 2) et's iid are gaussian
> 3) P(yt) is gaussian
> 4) No restrictions on P(Xt) and P(yt, Xt)

## Testing 3) using the Shapiro-Wilk test

The Shapiro-Wilk null hypothesis is that there is no difference between your distribution and a normal distribution.
  
```{r}
etf_names <- c("SPY" ,"XLI" ,"XLE" ,"XLK" ,"XLV", "XLU" ,"XLF" ,"XLY" ,"XLP" ,"XLB" ,"XHB", 
               "DBC" ,"USO" ,"UGA" ,"UNG" ,"SLV" ,"GLD", "TLT" ,"IEF" ,"IEI" ,"SHY" ,"MBB" ,
               "VCIT", "UUP" ,"FXE" ,"FXB" ,"FXY" ,"FXA" ,"RPI")
pval_threshold <- 0.05

shapiro_pvals <- list()
for (name in etf_names){
  stats <- shapiro.test(df[[name]])
  pval <- stats$p.value
  shapiro_pvals[[name]] <- pval 
}
shapiro_pvals_df <- do.call("rbind", shapiro_pvals) %>% as.data.table(keep.rownames=TRUE) %>%
  rename(pval=V1, etf=rn) %>% arrange(pval)

non_gaussian_etfs <- dim(shapiro_pvals_df[ shapiro_pvals_df$pval < pval_threshold, ])[1]
gaussian_etfs <- dim(shapiro_pvals_df[ shapiro_pvals_df$pval >= pval_threshold, ])[1]

print(paste0("Non-gaussian ETFs: ", non_gaussian_etfs, " Gaussian: ", gaussian_etfs))
```

## Running seqICP

```{r}
name <- etf_names[1]
Y <- df %>% select(one_of(name)) %>% as.matrix()
X <- df %>% select(-date, -one_of(etf_names)) %>% as.matrix()
seqICP(X=X,
       Y=Y,
       test = "hsic",
       par.test = list(grid = c(0,round(nrow(X)/2), nrow(X)),
                       complements = FALSE, link = sum,
                       alpha = 0.05, B = 100, permutation = FALSE),
       model = "iid",
       par.model = list(pknown = FALSE, p = 0, max.p = 10),
       max.parents = ncol(X),
       stopIfEmpty = TRUE,
       silent = TRUE)
```






