---
title: "VAR Simulations"
output:
---


```{r echo=FALSE, message=FALSE}
rm(list=ls())
library("seqICP")
library("data.table")
library("dplyr")
library("tsDyn")
```

### VAR(1) with no shifts

First, we will sample from a stationary VAR model with no shifts in distribution.

$$
x^a_{1,t} = 0.2x_{1,t-1} + 0.7x_{3,t-1} + N(0, 1)\\ 
x^a_{2,t} = 0.2x_{2,t-1} + N(0, 1)\\ 
x^a_{3,t} = 0.7x_{1,t-1} + 0.2x_{2,t-1} + N(0, 1)\\ 
$$
Thus, the SCM of $x_{1,t}$ is the same regardless of $e$ but there is a variance shift between the environments caused by a change in distribution in $x_{3,t-1}$.

Below we will sample from the above model:

```{r}

na = nb = 100

set.seed(1)
n <- na + nb

x1 <- matrix(data = NA, nrow = n, ncol = 1)
x2 <- matrix(data = NA, nrow = n, ncol = 1)
x3 <- matrix(data = NA, nrow = n, ncol = 1)

for (i in 1:n){
  
  # initialize all series
  if (i == 1){
    x1[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x2[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    next
  }
  
  x1[i,] <- 0.2*x1[i-1,] + 0.7*x3[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
  x2[i,] <- 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
  x3[i,] <- 0.7*x1[i-1,] + 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
}

var_sim <- cbind(x1, x2, x3)

var_sim_df <- var_sim %>% as.data.table()
colnames(var_sim_df) <- c("x1", "x2", "x3")
varnames <- colnames(var_sim_df)

write.csv(var_sim_df,
          file = file.path(getwd(), "data", "inputs", "gaussian_var_sim0.csv"),
          row.names = FALSE)

ts.plot(var_sim_df, type="l", col=c(1, 2, 3))
```

### Variance Shifted VAR(1)

To solve for the stationarity issue we had before, we will try to artificially introduce a variance shift into the VAR(1) process we defined before.

To do so, recall that $e \in \{a,b\}$ are two different environments. We define the following DGP for the variance shifted VAR(1) process:

$$
x^a_{1,t} = 0.2x^a_{1,t-1} + 0.7x^a_{3,t-1} + N(0, 1)\\ 
x^a_{2,t} = 0.2x^a_{2,t-1} + N(0, 1)\\ 
x^a_{3,t} = 0.7x^a_{1,t-1} + 0.2x^a_{2,t-1} + N(0, 1)\\ 
$$

$$
x^b_{1,t} = 0.2x^b_{1,t-1} + 0.7x^b_{3,t-1} + N(0, 1)\\ 
x^b_{2,t} = 0.2x^b_{2,t-1} + N(0, 1)\\ 
x^b_{3,t} = 0.7x^b_{1,t-1} + 0.2x^a_{2,t-1} + N(0, 3)\\ 
$$

Thus, the SCM of $x_{1,t}$ is the same regardless of $e$ but there is a variance shift between the environments caused by a change in distribution in $x_{3,t-1}$.

Below we will sample from the above model:

```{r}

na = nb = 100

set.seed(1)
n <- na + nb

x1 <- matrix(data = NA, nrow = n, ncol = 1)
x2 <- matrix(data = NA, nrow = n, ncol = 1)
x3 <- matrix(data = NA, nrow = n, ncol = 1)

for (i in 1:n){
  
  # initialize all series
  if (i == 1){
    x1[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x2[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    next
  }
  
  # sample time series iteratively conditioned on each environment
  if (i <= na){
    x1[i,] <- 0.2*x1[i-1,] + 0.7*x3[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x2[i,] <- 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- 0.7*x1[i-1,] + 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
  }
  else{
    x1[i,] <- 0.2*x1[i-1,] + 0.7*x3[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x2[i,] <- 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- 0.7*x1[i-1,] + 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 3)
  }
  
}

var_sim_shifted <- cbind(x1, x2, x3)

var_sim_shifted_df <- var_sim_shifted %>% as.data.table()
colnames(var_sim_shifted_df) <- c("x1", "x2", "x3")
varnames <- colnames(var_sim_shifted_df)

write.csv(var_sim_shifted_df,
          file = file.path(getwd(), "data", "inputs", "gaussian_var_sim1.csv"),
          row.names = FALSE)

ts.plot(var_sim_shifted_df, type="l", col=c(1, 2, 3))
```
### Variance Shifted SVAR(1) (Model C of section 6 from Pfister et al. (2018))

We hypothesize that the reason why seqICP was not able to find the correct set of causal parents is because the original DGP must have at least one contemporaneous relationship. This scenario is precisely the context of structural vector autoregressive models (SVAR).

In this context, we will try to artificially introduce a variance shift into the SVAR(1) process we defined before.

To do so, recall that $e \in \{a,b\}$ are two different environments. We define the following DGP for the variance shifted SVAR(1) process:

$$
x^a_{1,t} = 0.7x^a_{3,t} + 0.5x^a_{2,t-1} + N(0, 1)\\ 
x^a_{2,t} = 0.2x^a_{2,t-1} + N(0, 1)\\ 
x^a_{3,t} = 0.2x^a_{2,t} + N(0, 1)\\ 
$$

$$
x^b_{1,t} = 0.7x^b_{3,t} + 0.5x^a_{2,t-1} + N(0, 1)\\ 
x^b_{2,t} = 0.2x^b_{2,t-1} + N(0, 1)\\ 
x^b_{3,t} = 0.2x^a_{2,t} + N(0, 3)\\ 
$$

It can be seen by the above equations that $x_{1,t}$ depends on $x_{3,t}$. Below we simulate from this process:

```{r}

na = nb = 100

set.seed(1)
n <- na + nb

x1 <- matrix(data = NA, nrow = n, ncol = 1)
x2 <- matrix(data = NA, nrow = n, ncol = 1)
x3 <- matrix(data = NA, nrow = n, ncol = 1)

for (i in 1:n){
  
  # initialize all series
  if (i == 1){
    x1[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x2[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- rnorm(n = 1, mean = 0, sd = 1)
    next
  }
  
  # sample time series iteratively conditioned on each environment
  if (i <= na){
    x2[i,] <- 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- 0.2*x2[i,] + rnorm(n = 1, mean = 0, sd = 1)
    x1[i,] <- 0.7*x3[i,] + 0.5*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
  }
  else{
    x2[i,] <- 0.2*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
    x3[i,] <- 0.2*x2[i,] + rnorm(n = 1, mean = 0, sd = 3)
    x1[i,] <- 0.7*x3[i,] + 0.5*x2[i-1,] + rnorm(n = 1, mean = 0, sd = 1)
  }
  
}

var_sim_shifted2 <- cbind(x1, x2, x3)
var_sim_shifted_df2 <- var_sim_shifted2 %>% as.data.table()
colnames(var_sim_shifted_df2) <- c("x1", "x2", "x3")
varnames <- colnames(var_sim_shifted_df2)

write.csv(var_sim_shifted_df2,
          file = file.path(getwd(), "data", "inputs", "gaussian_var_sim2.csv"),
          row.names = FALSE)

ts.plot(var_sim_shifted2, type="l", col=c(1, 2, 3))
```
